Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-03-05 article.tex
Replace which <-> that

\begin{document}
\maketitle
\begin{abstract}
\Acf{CRN} play a pivotal role in diverse fields
such as systems biology, biochemistry, chemical engineering, and
epidemiology. High-level modelling of \acp{CRN} enables various simulation
approaches, including deterministic and stochastic methods. However,
existing Python tools for \ac{CRN} modelling typically wrap external C/C++
libraries for modelling and simulation, limiting their extensibility and
integration with the broader Python ecosystem. In response, we developed
Poincaré and SimBio, two novel Python packages for the definition and
simulation of dynamical systems and \acp{CRN}. Poincaré serves as a
foundation for dynamical system modelling, while SimBio extends this
functionality to \acp{CRN}, including support for the 
SBML. Poincaré and SimBio are developed as pure Python
packages enabling users to easily extend their simulation capabilities
by writing new or leveraging other Python packages. Moreover, this does
not compromise the performance, as code can be Just-In-Time compiled
with Numba. Our benchmark tests using curated models from the BioModels
repository, demonstrate that these tools may provide a potentially
superior performance advantage, compared to other existing tools.
Additionally, to ensure a user-friendly experience, our packages use
standard typed modern Python syntax that provides a seamless integration
with \acp{IDE}. Our Python-centric approach
significantly enhances code analysis, error detection, and refactoring
capabilities, positioning Poincaré and SimBio as valuable tools for the
modelling community.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

\Acp{CRN} are a fundamental concept of modelling
in numerous fields including systems biology, biochemistry, chemical
engineering and epidemiology. They are comprised of a set of chemical
species or biological entities
and a set of reactions that mediate transformations between them. These
systems can be modelled and simulated through multiple approaches:
deterministic \acp{ODE} to model
macroscopic behavior, \acp{SDE} to model
microscopic fluctuations, and jump processes (Gillespie-like
simulations) to account for the discreteness of populations. Instead of
directly writing the equations for each of these formulations which is
error-prone and difficult to reuse, these models can be written in a
higher-level description that can be compiled for these different types
of simulations.

Several tools already exist for defining and simulating \acp{CRN}.
BioSimulators.org (Shaikh et al. 2022), a registry of simulation tools,
list at least 15 Python software including COPASI (Hoops et al. 2006),
Tellurium (Choi et al. 2018) and PySB (Lopez et al. 2013). COPASI is a
standalone software with a \ac{GUI} for defining and simulating models. It's
widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow
advanced scripting. Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to
simulate models. PySB is a Python library that created a \ac{DSL} using
standard Python to define models which are then compiled to \acp{ODE} using a
Perl library called BioNetGen (L. A. Harris et al. 2016).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The variable name to which a component is assigned can be
  automatically saved in the component for introspection (i.e.,
  \texttt{Oscillator.x.name\ ==\ "x"}),
\item
  It provides a namespace that allows to easily define multiple
  independent models in the same script,
\item
  It allows \acp{IDE} to provide autocomplete and refactoring capabilities
  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x},
  \texttt{v} and \texttt{eq}),
\item
  It allows creation of instances which can be composed into a bigger
  model (Figure~\ref{fig-composition}).
\end{enumerate}

@@

For this last point, \acp{IDE} that support \texttt{dataclass\_transform} (De
Bonte and Traut 2021) can provide a tooltip with the expected signature
(Figure~\ref{fig-ide}). This requires the use of type
annotations which play a more significant role in static type checking,
as they can help to identify errors before running the code. For
instance, to parameterize the initial conditions of variables we have to
use a \texttt{Constant}. If we try to use a \texttt{Parameter}, which
could be a time-dependent expression, it is flagged as a type error
(Figure~\ref{fig-ide}).

To simulate a system, we created a \texttt{Simulator} instance
(Figure~\ref{fig-sim}), which compiles the \ac{RHS} of a given system and interfaces
with \ac{ODE} solvers. By default, it creates a first-order
\ac{ODE} system using \texttt{numpy} as a backend. This can be
easily switched to other solvers. The Simulator wraps the output in a
\texttt{pandas.DataFrame}, which can be easily plotted with the standard
\texttt{plot} method.

Switching backends to \texttt{"numba"} can improve model runtime by a
factor up to x30 or more for big or long running models, however incurs
in a compile time penalty for the first run which must be taken into
account.

@@

For the reaction networks, our focus is on first-order differential
equations that describe the rate of change of species. SimBio simplifies
the definition of these network models by introducing \texttt{Species},
which consists of a \texttt{poincare.Variable} and a stoichiometric
number, and \texttt{RateLaw}, a construct that converts reactants into
products taking into account the stoichiometry
(Figure~\ref{fig-simbio}). Additionally, SimBio features
\texttt{MassAction}, a subclass of \texttt{RateLaw}, which intuitively
incorporates reactants into the rate law (Figure~\ref{fig-simbio}).

@@

To evaluate SimBio's reproducibility, we analyzed curated \ac{SBML} models
from BioModels (Malik-Sheriff et al. 2020). Among the first 100 curated
models from BioModels we selected 60 which did not contains events, as
SimBio doesn't support events yet. We simulated the selected models with
COPASI and used the simulated results as ground truth, and demonstrated
that SimBio reproduces the results with minor numerical differences
depending on the solver tolerance.

@@

For COPASI and Tellurium/RoadRunner, we noted that its runtime depended
on the number of evaluation points, something that does not seem to
happen with SimBio (Figure~\ref{fig-runtime}, left). While SimBio's
NumPy backend is slower than both COPASI and RoadRunner, we obtained an
order of magnitude speed-up using the numba backend putting it on par
with them. A user might have to consider the trade-off between
compilation and run times, as the compilation of the
\ac{RHS} code might take longer than the runtime itself, and not be worth
it for running the model only once. Another speed-up in the runtime can
be had by switching the LSODA \texttt{scipy} solver for a more efficient
\texttt{numbalsoda} implementation, which avoids the Python interpreter
between each of the integration steps. This last combination beats all
other methods, which is also true for the other models we tested
(Figure~\ref{fig-runtime}, right).

>>>>

\begin{document}
\maketitle
\begin{abstract}
\Acf{CRN} play a pivotal role in diverse fields
such as systems biology, biochemistry, chemical engineering, and
epidemiology. High-level modelling of \acp{CRN} enables various simulation
approaches, including deterministic and stochastic methods. However,
existing Python tools for \ac{CRN} modelling typically wrap external C/C++
libraries for modelling and simulation, limiting their extensibility and
integration with the broader Python ecosystem. In response, we developed
Poincaré and SimBio, two novel Python packages for the definition and
simulation of dynamical systems and \acp{CRN}. Poincaré serves as a
foundation for dynamical system modelling, while SimBio extends this
functionality to \acp{CRN}, including support for the 
SBML. Poincaré and SimBio are developed as pure Python
packages enabling users to easily extend their simulation capabilities
by writing new or leveraging other Python packages. Moreover, this does
not compromise the performance, as code can be Just-In-Time compiled
with Numba. Our benchmark tests using curated models from the BioModels
repository demonstrate that these tools may provide a potentially
superior performance advantage compared to other existing tools.
Additionally, to ensure a user-friendly experience, our packages use
standard typed modern Python syntax that provides a seamless integration
with \acp{IDE}. Our Python-centric approach
significantly enhances code analysis, error detection, and refactoring
capabilities, positioning Poincaré and SimBio as valuable tools for the
modelling community.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

\Acp{CRN} are a fundamental concept of modelling
in numerous fields including systems biology, biochemistry, chemical
engineering and epidemiology. They are comprised of a set of chemical
species or biological entities
and a set of reactions that mediate transformations between them. These
systems can be modelled and simulated through multiple approaches:
deterministic \acp{ODE} to model
macroscopic behavior, \acp{SDE} to model
microscopic fluctuations, and jump processes (Gillespie-like
simulations) to account for the discreteness of populations. Instead of
directly writing the equations for each of these formulations, which is
error-prone and difficult to reuse, these models can be written in a
higher-level description that can be compiled for these different types
of simulations.

Several tools already exist for defining and simulating \acp{CRN}.
BioSimulators.org (Shaikh et al. 2022), a registry of simulation tools,
list at least 15 Python software including COPASI (Hoops et al. 2006),
Tellurium (Choi et al. 2018) and PySB (Lopez et al. 2013). COPASI is a
standalone software with a \ac{GUI} for defining and simulating models. It's
widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow
advanced scripting. Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to
simulate models. PySB is a Python library that created a \ac{DSL} using
standard Python to define models, which are then compiled to \acp{ODE} using a
Perl library called BioNetGen (L. A. Harris et al. 2016).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The variable name to which a component is assigned can be
  automatically saved in the component for introspection (i.e.,
  \texttt{Oscillator.x.name\ ==\ "x"}),
\item
  It provides a namespace that allows to easily define multiple
  independent models in the same script,
\item
  It allows \acp{IDE} to provide autocomplete and refactoring capabilities
  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x},
  \texttt{v} and \texttt{eq}),
\item
  It allows creation of instances that can be composed into a bigger
  model (Figure~\ref{fig-composition}).
\end{enumerate}

@@

For this last point, \acp{IDE} that support \texttt{dataclass\_transform} (De
Bonte and Traut 2021) can provide a tooltip with the expected signature
(Figure~\ref{fig-ide}). This requires the use of type
annotations, which play a more significant role in static type checking
as they can help to identify errors before running the code. For
instance, to parameterize the initial conditions of variables we have to
use a \texttt{Constant}. If we try to use a \texttt{Parameter}, which
could be a time-dependent expression, it is flagged as a type error
(Figure~\ref{fig-ide}).

To simulate a system, we created a \texttt{Simulator} instance
(Figure~\ref{fig-sim}) that compiles the \ac{RHS} of a given system and interfaces
with \ac{ODE} solvers. By default, it creates a first-order
\ac{ODE} system using \texttt{numpy} as a backend. This can be
easily switched to other solvers. The Simulator wraps the output in a
\texttt{pandas.DataFrame}, which can be easily plotted with the standard
\texttt{plot} method.

Switching backends to \texttt{"numba"} can improve model runtime by a
factor up to x30 or more for big or long running models, however incurs
in a compile time penalty for the first run that must be taken into
account.

@@

For the reaction networks, our focus is on first-order differential
equations that describe the rate of change of species. SimBio simplifies
the definition of these network models by introducing \texttt{Species},
which consists of a \texttt{poincare.Variable} and a stoichiometric
number, and \texttt{RateLaw}, a construct that converts reactants into
products taking into account the stoichiometry
(Figure~\ref{fig-simbio}). Additionally, SimBio features
\texttt{MassAction}, a subclass of \texttt{RateLaw}, that intuitively
incorporates reactants into the rate law (Figure~\ref{fig-simbio}).

@@

To evaluate SimBio's reproducibility, we analyzed curated \ac{SBML} models
from BioModels (Malik-Sheriff et al. 2020). Among the first 100 curated
models from BioModels we selected 60 that did not contains events, as
SimBio doesn't support events yet. We simulated the selected models with
COPASI and used the simulated results as ground truth, and demonstrated
that SimBio reproduces the results with minor numerical differences
depending on the solver tolerance.

@@

For COPASI and Tellurium/RoadRunner, we noted that its runtime depended
on the number of evaluation points, something that does not seem to
happen with SimBio (Figure~\ref{fig-runtime}, left). While SimBio's
NumPy backend is slower than both COPASI and RoadRunner, we obtained an
order of magnitude speed-up using the numba backend putting it on par
with them. A user might have to consider the trade-off between
compilation and run times, as the compilation of the
\ac{RHS} code might take longer than the runtime itself, and not be worth
it for running the model only once. Another speed-up in the runtime can
be had by switching the LSODA \texttt{scipy} solver for a more efficient
\texttt{numbalsoda} implementation that avoids the Python interpreter
between each of the integration steps. This last combination beats all
other methods, which is also true for the other models we tested
(Figure~\ref{fig-runtime}, right).
